Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-10-01 docs/papers/shallow_performance/shallow_perf.tex
forgot to add a \ to the end of \psykal in the paper so any spaces were removed.

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

In common with OP2, the \psykal approach separates out the science
code and the performance related code and does this via a separate
layer. The calls that specify parallelism in both approaches are
similar in terms of where they are placed in the code and in their
semantics but the \psykal approach supports the specification of more
than one Kernel in a parallel region of code, compared with one for
OP2, giving more scope for optimisation. Another difference is that
the description metadata is provided with the Kernel in the \psykal
approach whereas it is provided as part of the Kernel call in OP2.

While the \psykal approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.

>>>>

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique which we call \psykal\ to finite difference Ocean models. 
%
In \psykal\, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal\ version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion {\it etc.}) are then
applied to the \psykal\ version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem
sizes. The only exception to this (other than where we see performance
improvement) is the Gnu compiler on Haswell for one problem size. Our
tests indicate that this may be due to immature support for that
architecture in the Gnu compiler -- no such problem is seen on the Ivy
Bridge system. Further, the original code performed poorly using the
IBM compiler on Power8 and needed to be modified to obtain performant
code.
%%
Therefore, the \psykal\ approach can be used with negligible
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the \psykal\ version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

In common with OP2, the \psykal\ approach separates out the science
code and the performance related code and does this via a separate
layer. The calls that specify parallelism in both approaches are
similar in terms of where they are placed in the code and in their
semantics but the \psykal\ approach supports the specification of more
than one Kernel in a parallel region of code, compared with one for
OP2, giving more scope for optimisation. Another difference is that
the description metadata is provided with the Kernel in the \psykal\
approach whereas it is provided as part of the Kernel call in OP2.

While the \psykal\ approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.
